---
title: "DSCC/CSC/TCS 462 Assignment 0"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
subtitle: Due Thursday, September 8, 2022 by 3:59 p.m.
fontsize: 12pt
---
```{r, echo=FALSE}
knitr::opts_chunk$set(error = TRUE)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T,tidy=TRUE, tidy.opts=list(width.cutoff=
                                                                   60))
```
  
This assignment will cover material from Lectures 1 and 2. You are expected to 
use the `ggplot2` library in `R` for completing all the graphics. To learn more 
about graphics using `ggplot2`, please read through the guide available here:   http://www.cookbook-r.com/Graphs/. This is a wonderful open source textbook that
walks through examples of many different graphics in `ggplot2`. If you have not
done so already, start by installing the library. In the R console (i.e. NOT in
your .RMD file), run the code `install.packages("ggplot2")`. Then, in your .RMD
file, load the library as follows:
```{r, echo=T}
library(ggplot2)
```

For this first assignment, we will use the "car_sales.csv" dataset, which
includes information about 152 different cars. In particular, we will mainly
focus on the selling price of cars throughout this assignment.


1. Getting familiar with the dataset via exploratory data analysis. 
    a. Read the data into RStudio and summarize the data with the `summary()` 
    function. 
```{r, echo=T, warning=FALSE}
library(ggplot2)
car_sales <- read.csv("car_sales.csv", header=TRUE, sep = ",")
summary(car_sales)
```    
    
    b. How many bins does Sturges' formula suggest we use for a histogram of 
    `price`? Show your work. 
```{r, echo=T, warning=FALSE}
n = length(car_sales$price) # number of rows in price column
n_bins = ceiling((log2(n)) + 1)
paste("Number of bins = ", n_bins, ", as per Sturge's formula")
```      
    c. Create a histogram of `price` using the number of bins suggested by 
    Sturges' formula in 1b. Make sure to appropriately title the histogram and 
    label the axes. Comment on the center, shape, and spread. 
```{r, echo=T, warning=FALSE}
# Histogram
ggplot(car_sales, aes(x=price)) + geom_histogram(color="black", fill="white",
                    bins = n_bins) + ggtitle("Histogram of Car Prices Data") + 
 labs(x="Price", y="Frequency") +  theme(plot.title = element_text(hjust = 0.5))
# Comment on center, shape and spread
# It is a unimodal right skewed distribution, and is asymmetrical.
```   

    
2. Measures of center and spread for the selling price of cars.
    a. Calculate the mean, median, and 10\% trimmed mean of the selling price. 
    Report the mean, median, and 10\% trimmed mean on the histogram. 
    In particular, create a red vertical line on the histogram at the mean, 
    and report the value of the mean in red next to the line using the form 
    "$\bar{x}=$". Create a blue vertical line on the histogram at the median,
    and report the value of the median in blue next to the line using the form 
    "$\tilde{x}=$". Create a green vertical line on the histogram at the 10\% 
    trimmed mean, and report the value of the 10\% trimmed mean in green next 
    to the line using the form "$\bar{x}_{10}=$" (to get $\bar{x}_{10}$ to 
    print on the plot, use `bar(x)[10]` within the `paste()` function). 
```{r, echo=T, warning=FALSE}
price = car_sales$price 
mean_value = mean(price)
med_val = median(price)
trimmed_mean = mean(price, trim = 0.1)
ggplot(car_sales, aes(x=price)) + geom_histogram(color="black", fill="white",
       bins = n_bins) + ggtitle("Histogram of Car Prices Data") + 
  geom_vline(xintercept = mean_value, col = "red") +
  annotate("text", x=mean_value + 6200, y=60,
           label = substitute(paste(bar(x), " = ", m),
                              list(m=mean_value)), col="red")+
  geom_vline(xintercept = med_val, col = "blue") +
  annotate("text", x=med_val + 5000, y=50,
           label = substitute(paste(tilde(x), " = ", m),
                              list(m=med_val)), col="blue")+
  geom_vline(xintercept = trimmed_mean, col = "green") +
  annotate("text", x=trimmed_mean + 7000, y=40,
           label = substitute(paste(bar(x)[10], " = ", m),
                              list(m=trimmed_mean)), col="green") + 
          labs(x="Car Prices", y="Frequency") +
  theme(plot.title = element_text(hjust = 0.5))
```         
    
    b. Calculate and report the 25th and 75th percentiles. 
```{r, echo=T, warning=FALSE}
q1 = unname(quantile(price, 0.25)) 
q3 = unname(quantile(price, 0.75))
paste("25th percentile is", q1, "and 75th percentile is", q3)
```         
    c. Calculate and report the interquartile range. 
```{r, echo=T, warning=FALSE}
iqr = IQR(price)
paste("Interquartile Range is", iqr)
first_quartile = quantile(price, 0.25)
third_quartile = quantile(price, 0.75)
```         
    d. Calculate and report the standard span, the lower fence, and the upper 
    fence.
    ```{r, echo=T, warning=FALSE}
Standard_Span = 1.5*iqr
lower_fence = unname(first_quartile - Standard_Span)
upper_fence = unname(third_quartile + Standard_Span)

paste("Standard Span is", Standard_Span)
paste("Lower Fence is", lower_fence, "and Upper Fence is", upper_fence)
```       
    e. Are there any outliers? Subset the outlying points. Use code based on 
    the following:
    ```{r echo=T}
    car_sales[car_sales$price >= upper_fence, ] #upper outliers
    car_sales[car_sales$price <= lower_fence, ] #lower outliers 
    # Use upper and lower fence values from part g.
    car_sales[car_sales$price >= upper_fence, ] -> upper_outliers
    car_sales[car_sales$price >= lower_fence, ] -> lower_outliers
    ```
    
    f. Calculate and report the variance, standard deviation, and coefficient 
    of variation of car prices. 
```{r, echo=T, warning=FALSE}
variance.value = var(car_sales$price)
sd.value = sd(car_sales$price)
coeff.variation = sd.value / mean_value

paste("Variance is", variance.value)
paste("Standard Deviation is", sd.value)
paste("Coefficient of Variation is", coeff.variation)
```        
    g. We have seen from the histogram that the data are skewed. Calculate and
    report the skewness. Comment on this value and how it matches with what you
    visually see in the histogram. 
    ```{r, echo=T, warning=FALSE}
install.packages("moments")
library("moments") # used moments library to calculate skewness
paste("Skewness is",skewness(price))
# comment
# Since the skewness is positive, it indicates that our distribution is right
# skewed (positive skew)
``` 
3. Transforming the data.
    a. Use a Box-Cox power transformation to appropriately transform the data. 
    In particular, use the `boxcox()` function in the `MASS` library. Report 
    the recommended transformation. Do not apply this transformation to the data
    yet. (Note: the `boxcox` function automatically produces a plot. You do NOT 
    need to make this in `ggplot2`.)
```{r, echo=T, warning=FALSE}
library('MASS')
bc.plot = boxcox(car_sales$price~1)
```         
    b. Apply the exact Box-Cox recommended transformation (rounded to four 
    decimal places) to the data (this transformation is hereon referred to as 
    the Box-Cox transformed data). Use the `summary()` function to summarize the results of this transformation. 
```{r, echo=T, warning=FALSE}
lambda = bc.plot$x[bc.plot$y==max(bc.plot$y)] 
transformed_data = (car_sales$price ^ round(lambda, 4) - 1) / round(lambda, 4)
print("Summary of BoxCox Transformed Data")
summary(transformed_data)

```         
    c. Create a histogram of the Box-Cox transformed data using the number of 
    bins suggested by Sturges' formula. On this histogram, report the mean, 
    median, and 10\% trimmed mean using the same formatting options as in part 
    2a above. Comment on the center, shape, and spread. 
```{r, echo=T, warning=FALSE}
    
td.mean_value = mean(transformed_data)
td.med_val = median(transformed_data)
td.trimmed_mean = mean(transformed_data, trim = 0.1)
    ggplot(car_sales, aes(x=transformed_data)) + geom_histogram(color="black", 
                                                                fill="white",
         bins = n_bins) + ggtitle("Histogram of Car Prices Data") + 
  geom_vline(xintercept = td.mean_value, col = "red") +
  annotate("text", x=td.mean_value + 0.0020, y=60,
           label = substitute(paste(bar(x), " = ", m),
                              list(m=td.mean_value)), col="red")+
  geom_vline(xintercept = td.med_val, col = "blue") +
  annotate("text", x=td.med_val + 0.0025, y=50,
           label = substitute(paste(tilde(x), " = ", m),
                              list(m=td.med_val)), col="blue")+
  geom_vline(xintercept = td.trimmed_mean, col = "green") +
  annotate("text", x=td.trimmed_mean + 0.003, y=40,
           label = substitute(paste(bar(x)[10], " = ", m),
           list(m=td.trimmed_mean)), col="green") + 
           labs(x="Car Prices - BoxCox", y="Frequency") +
           theme(plot.title = element_text(hjust = 0.5))
# comment: 
# After boxcox transformation it has become a unimodal normal distribution, 
# and is almost symmetrical, and there are no potential outliers.  
```         
    d. As an alternative to the Box-Cox transformation, let's also use a log 
    transformation. Apply the log transformation to the original `price` data 
    (this transformation is hereon referred to as the log transformed data). 
    Use the `summary()` function to summarize the results of this transformation. 
```{r, echo=T, warning=FALSE}

log_td = log(car_sales$price)
print("Summary of Log Transformed Data")
summary(log_td)
```        
    e. Create a histogram of the log transformed data using the number of bins 
    suggested by Sturges' formula. On this histogram, report the mean, median, 
    and 10\% trimmed mean using the same formatting options as in part 2a and 3c
    above. Comment on the center shape and spread. 
```{r, echo=T, warning=FALSE}


log.mean_value = mean(log_td)
log.med_val = median(log_td)
log.trimmed_mean = mean(log_td, trim = 0.1)
ggplot(car_sales, aes(x=log_td)) + geom_histogram(color="black", fill="white",
       bins = n_bins) + ggtitle("Histogram of Car Prices Data") + 
      geom_vline(xintercept = log.mean_value, col = "red") +
  annotate("text", x=log.mean_value + 0.15, y=40,
      label = substitute(paste(bar(x), " = ", m),
      list(m=log.mean_value)), col="red")+
  geom_vline(xintercept = log.med_val, col = "blue") +
  annotate("text", x=log.med_val + 0.15, y=37,
           label = substitute(paste(tilde(x), " = ", m),
                              list(m=log.med_val)), col="blue")+
  geom_vline(xintercept = log.trimmed_mean, col = "green") +
  annotate("text", x=log.trimmed_mean + 0.15, y=34,
           label = substitute(paste(bar(x)[10], " = ", m),
                              list(m=log.trimmed_mean)), col="green") + 
  labs(x="Car Prices - Log", y="Frequency") +
  theme(plot.title = element_text(hjust = 0.5))
# Comment
# It is a unimodal normal distribution, and is almost symmetrical, and there are
# no potential outliers, but it is slightly skewed to the right.
```        
    f. Create a qqplot for the original data, a qqplot for the Box-Cox 
    transformed  data, and a qqplot of the log transformed data. 
    Comment on the results. 
```{r, echo=T, warning=FALSE}
qqnorm(price, main = "Q-Q Plot of Original Given Data")
qqline(price, col = "red", lwd = 2)

qqnorm(transformed_data, main = "Q-Q Plot of BoxCox Transformed Data")
qqline(transformed_data, col = "blue", lwd = 2)

qqnorm(log_td, main = "Q-Q Plot of Log Transformed Data")
qqline(log_td, col = "green", lwd = 2)

# For original data: The data points vary largely from the theoretical line which
# indicates skewness.
# For BoxCox transformation: The data points almost fall on the theoretical line
# which follows a normal distribution
# For log transformation: The data points fall mostly fall on the line but deviates
# a little when compared to the boxcox data plot.default
# Therefore, boxcox transformed data is almost like a normal distribution.

```         
    g. Evaluate the empirical rule for the original data, the Box-Cox 
    transformed data, and the log transformed data. In particular, make a table 
    similar to that on slide 71 of the Chapter 2 notes. Comment on the results. 
    Do either of the transformed data seem to be "better" to work with? Note, 
    you can use code similar to the following to answer this question:
```{r echo=TRUE}
mat <- matrix(NA, nrow=9, ncol=5)

### Set row names and column names
rownames(mat) <- c("Original","","","Box-Cox","","","Log","","")
colnames(mat) <- c("x","xbar-k*s", "xbar+k*s", "Theoretical %","Actual %")

### Fill in known quantities
mat[,1] <- c(1,2,3)
mat[,4]<-c(68,95,99.7)
# for xbar - k*s:
mat[1,2] <- mean(price)-1*sd(price)
mat[2,2] <- mean(price)-2*sd(price)
mat[3,2] <- mean(price)-3*sd(price)

mat[4,2] <- mean(transformed_data)-1*sd(transformed_data)
mat[5,2] <- mean(transformed_data)-2*sd(transformed_data)
mat[6,2] <- mean(transformed_data)-3*sd(transformed_data)

mat[7,2] <- mean(log_td)-1*sd(log_td)
mat[8,2] <- mean(log_td)-2*sd(log_td)
mat[9,2] <- mean(log_td)-3*sd(log_td)

# for xbar + k*s:
mat[1,3] <- mean(price)+1*sd(price)
mat[2,3] <- mean(price)+2*sd(price)
mat[3,3] <- mean(price)+3*sd(price)

mat[4,3] <- mean(transformed_data)+1*sd(transformed_data)
mat[5,3] <- mean(transformed_data)+2*sd(transformed_data)
mat[6,3] <- mean(transformed_data)+3*sd(transformed_data)

mat[7,3] <- mean(log_td)+1*sd(log_td)
mat[8,3] <- mean(log_td)+2*sd(log_td)
mat[9,3] <- mean(log_td)+3*sd(log_td)

# for Actual %:
mat[1,5] <- sum(price >=mean(price)-1*sd(price) 
                & price <= mean(price)+1*sd(price))/length(price)*100
mat[2,5] <- sum(price >=mean(price)-2*sd(price) 
                & price <= mean(price)+2*sd(price))/length(price)*100
mat[3,5] <- sum(price >=mean(price)-3*sd(price) 
                & price <= mean(price)+3*sd(price))/length(price)*100

mat[4,5] <- sum(transformed_data >=mean(transformed_data)-1*sd(transformed_data) 
                & transformed_data <= mean(transformed_data)+1*sd(transformed_data))/length(transformed_data)*100
mat[5,5] <- sum(transformed_data >=mean(transformed_data)-2*sd(transformed_data) 
                & transformed_data <= mean(transformed_data)+2*sd(transformed_data))/length(transformed_data)*100
mat[6,5] <- sum(transformed_data >=mean(transformed_data)-3*sd(transformed_data) 
                & transformed_data <= mean(transformed_data)+3*sd(transformed_data))/length(transformed_data)*100

mat[7,5] <- sum(log_td >=mean(log_td)-1*sd(log_td) 
                & log_td <= mean(log_td)+1*sd(log_td))/length(log_td)*100
mat[8,5] <- sum(log_td >=mean(log_td)-2*sd(log_td) 
                & log_td <= mean(log_td)+2*sd(log_td))/length(log_td)*100
mat[9,5] <- sum(log_td >=mean(log_td)-3*sd(log_td) 
                & log_td <= mean(log_td)+3*sd(log_td))/length(log_td)*100


### Create a table
library(knitr)
kable(x=mat, digits=2,row.names=T, format="markdown")
# comment:
# In actual data, the theoretical value and the actual value varies a lot.
# In boxcox transformed data, the actual value is almost closer to the theoretical 
# value, and is much better compared to the log transformed data. So, boxcox 
# transformation is much better for our data

```
    h. In your own words, provide some intuition about (1) why car price may 
    not follow a normal distribution, and (2) why it may be useful to transform
    the data into a form that more closely follows a normal distribution.
```{r, echo=T, warning=FALSE}
# (1) The reason for car prices not following normal distribution is that, not all 
# people prefer expensive cars and if the prices are high, the number of people 
# buying it will obviously be lesser and more number of people prefer buying 
# mid-range and less expensive cars, and that is why, our data had its peak
# towards the left of mean, and most our data is right skewed.

# (2) If it follows normal distribution, all of the mean, median and mode would be
# the same, and also with just the mean and standard deviation, we can get 
# insights about the whole distribution.
```         
    
Short Answers:

  * About how long did this assignment take you? Did you feel it was too long, 
  too short, or reasonable? 
Answer: It took me quite some time, as I have never used R before, and I felt 
it was longer and a little tougher for me.
  
  * Who, if anyone, did you work with on this assignment?
Answer: I was not able to solve every question by myself, so I discussed with my
friend Ajeesh Ajayan to get some insights on few of the questions and then 
solved the problems.
  
  * What questions do you have relating to any of the material we have covered 
  so far in class?    
Answer: I am quite clear with the concepts taught so far.